A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>
edits^3

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    % NSP-22NOV: "unary node" and "ARG" are not defined so I tried to provide better context here
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees where they
    are not coalescence events.
    We show how these non-coalescing haplotypes (``unary nodes'') may be
    inserted into the ancestral recombination graph (ARG), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting inferred ARGs are smaller, faster to compute with,
    and contain substantially more information about inheritance % NSP-22NOV: "substantially more information" in what regard?
    than inferred ARGs without unary nodes.
    We provide efficient algorithms to infer unary nodes within existing ARGs,
    new metrics of agreement and disagreement between inferred ARGs,
    and explore some consequences of unary nodes for ARGs inferred from real data.
\end{abstract}

@@

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
% NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
% $\mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]$
% and also in next equation for consistency.
% HF24-NOV: I think I disagree but we can get a third opinion.
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k,\ n_1) = S(T^{(2)}_k,\ n_2)}  
    % NSP-22NOV: Added a missing comma. Also the comma at the very end looks like a prime, which is confusing, so I'm removing grammatical punctuation from these non-inline equations.
    % HF-24NOV: Noted
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} 
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) 
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) 
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) 
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
% NSP-22NOV: as before I'd suggest using \mathbb{I}[...] for the indicator function
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k,\ n_2) = s(T^{(1)}_k,\ n)} 
\end{align*}
the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could possibly be defined
using subgraph-prune-and-regraft moves \citet{deng2024robust}.

@@

\paragraph{Metrics on ARGs}
The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
This is by design: in practice it is not possible to infer all aspects
of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
and so we wanted to quantify
``How much of the true relationships does this ARG represent?''
However, it is worth noting that the symmetrized version
$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
and view each ARG as a subset of the space $[0,L) \times N \times N$,
where $N$ is the shared set of nodes.
Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
$|\T_1 \setminus \T_2|$,
and so the symmetrized version is the measure of the symmetric difference
$|\T_1\Delta\T_2|$,
which is well-known to be a metric. % NSP-22NOV: "well-known" ==> give a citation here
% HF-24NOV: Peter would we cite Rudin?
If the two ARGs have the same number of nodes,
we can consider all bijections between their nodes.
The symmetric difference between $\T_1$ and $\T_2$,
related through each bijection is a metric.
Then the minimum over all such metrics will still be a metric
since the minimum over a finite number of metrics is also a metric.
 % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
 % HF-24NOV: Hopefully it is clearer now
This also extends to two tree sequences with different numbers of nodes, $|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.

@@

%\begin{figure}
%    \caption{
%        Supplementary figure: runtime of discrepancy function as a function of number of trees
%        (which we modulate by changing sequence length).
%        \label{fig:speed_discrepancy}
%    }
%\end{figure}

>>>>

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    % NSP-22NOV: "unary node" and "ARG" are not defined so I tried to provide better context here
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees where they
    are not coalescence events.
    We show how these non-coalescing haplotypes (``unary regions'') may be
    inserted into the ancestral recombination graph (ARG), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting inferred ARGs are smaller, faster to compute with,
    and contain substantially more information about inheritance % NSP-22NOV: "substantially more information" in what regard?
    than inferred ARGs without unary nodes.
    We provide efficient algorithms to infer unary nodes within existing ARGs,
    new metrics of agreement and disagreement between inferred ARGs,
    and explore some consequences of unary nodes for ARGs inferred from real data.
\end{abstract}

@@

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
	\mathcal{M}(\T_1, n1; \T_2, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
\end{align*}
to be the subset of all places $k$ where $n_1$ and $n_2$ share the same sample set in $T_k^{(1)}$ and $T_k^{(2)}$, respectively.
Further, we define
% NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
% $\mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]$
% and also in next equation for consistency.
% HF24-NOV: I think I disagree but we can get a third opinion.
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k\in \mathcal{M}(\T_1, n1; \T_2, n2)} (a_k - a_{k-1}) ,  
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
% NSP-22NOV: as before I'd suggest using \mathbb{I}[...] for the indicator function
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k,\ n_2) = s(T^{(1)}_k,\ n)} 
\end{align*}
as the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
A similar metric on tree sequences could possibly be defined
using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.

@@

\paragraph{Metrics on ARGs}
The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
This is by design: in practice it is not possible to infer all aspects
of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
and so we wanted to quantify
``How much of the true relationships does this ARG represent?''
However, it is worth noting that the symmetrized version
$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
and view each ARG as a subset of the space $[0,L) \times N \times N$,
where $N$ is the shared set of nodes.
Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
$|\T_1 \setminus \T_2|$,
and so the symmetrized version is the measure of the symmetric difference
$|\T_1\Delta\T_2|$,
which is well-known to be a metric \citep{rudin1976principles}. 
If the two ARGs have the same number of nodes,
we can consider all bijections between their nodes.
The symmetric difference between $\T_1$ and $\T_2$,
related through each bijection is a metric.
Then the minimum over all such metrics will still be a metric
since the minimum over a finite number of metrics is also a metric.
 % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
 % HF-24NOV: Hopefully it is clearer now
This also extends to two tree sequences with different numbers of nodes, $|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.

@@
